---
title: "Data Manipulation with dplyr"
author: "Rae Spriggs"
date: "2022-09-22"
output: github_document
---

Once you've imported data, you're going to need to do some cleaning up. 

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  fig.width = 6,
  fig.asp = 6, 
  out.width = "90%"
)
```

```{r}
library(tidyverse)
options(tibble.print_min = 3)

litters_data = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)
```

# Ways to SELECT variables (entire columns)

```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```

Select variables in order left to right on the tibble
```{r}
select(litters_data, group:gd_of_birth)
```

Select by telling it to remove certain variables
```{r}
select(litters_data, -pups_survive)

select(litters_data, -pups_survive, -group)
```

## Ways to rename variables 
 
```{r}
select(litters_data, GROUP = group, LITTER_NUMBER = litter_number)

##OR - another way to do the same thing

rename(litters_data, GROUP = group, LITTER_NUMBER = litter_number)
```

## Helper functions (starts with, ends with, and contains)

```{r}
select(litters_data, starts_with("gd"))
```

```{r}
select(litters_data, litter_number, pups_survive, everything())
```

```{r}
relocate(litters_data, litter_number, pups_survive)
```

# Ways to FILTER observations (entire rows)

You will often filter using comparison operators (>, >=, <, <=, ==, and !=). You may also use %in% to detect if values appear in a set, and is.na() to find missing values. The results of comparisons are logical – the statement is TRUE or FALSE depending on the values you compare – and can be combined with other comparisons using the logical operators & and |, or negated using !.

Filtering certain observations by logical criteria 
```{r}
filter(litters_data, 
gd_of_birth == 20,
pups_born_alive >= 2,
pups_survive != 4,
!(pups_survive == 4))

```

```{r}
filter(litters_data, group %in% c("Con7", "Con8"))
```

```{r}
filter(litters_data, group == "Con7" & gd_of_birth == 20)
```

## Drop NA function

```{r}
library(tidyr)
drop_na(litters_data) 
```

# Ways to MUTATE data (change or create new variables)

```{r}
mutate(litters_data,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```

# Ways to ARRANGE data (sorting rows in data)

```{r}
head(arrange(litters_data, desc(group), pups_born_alive), 10)

## head function lets you print the first 10 observations
## desc is descending for the variable it's applied to 
```

# PIPING / %>% 

Shorthand is CMD + SHIFT + M for %>%
```{r}
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)

litters_data
```

